The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-12-14 script/en/chapters/primes.tex
footnote fixes
CT script: Minor changes (updated intro)
CT script: Minor formal changes in chapter 3

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Nov. 2012)

@@

% --------------------------------------------------------------------------
\section{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Number!natural} numbers $\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

\noindent Prime numbers are particularly important for three reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorized in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors not equal 1 are called 
\index{Number!composite} {\em composite} numbers. 
These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

The following table contains the current record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \url{http://primes.utm.edu/largest.html}
and at
     \url{http://primes.utm.edu/mersenne/index.html}.
}:

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2012)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

When the project started in 2002 there have been 17 possible candidates $< 78557$ (this is the reason for the project's name ``Seventeen or Bust''). It is sufficient to find one single counter-example, to exclude a candidate k, which means to find a single $n>=1$, where $N = k \cdot 2^{n}+1$ is prime. So it is only a byproduct of this task that this also generates new monster primes.
} 
\footnotetext{%
Generalized Fermat number: $ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $
\index{Fermat!number!generalized}  } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{Factorization} (expression as a product of prime factors)\footnote{%
  Using CT1\index{CrypTool 1} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CT1 can factorize with the quadratic sieve (QS) on a single PC in a reasonable
  time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CT1. \\
  With CT2 you can use both the QS and the GNFS (general number field sieve),
  and CT2 can distribute the calculations on several computers.\\
  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{Factorization!factoring records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +10 pt
For the first 41            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 47th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

@@

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \url{http://www.mersenne.org/history.htm}.
}$^,$\footnote{%
Always, when a new record is published in the respective forums the same
and often ironic discussions start: Does this kind of research have a deeper
sense? Can this result be applied for anything useful?
The answer is, that we don't know it yet. In fundamental research one
cannot see at once whether and how it brings mankind forward.
%But this is the normal case with basic research, that one cannot see at once, 
}:

@@

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of July 2009)}
\end{center}
\end{table}

@@

Dr. Richard Crandall\index{Crandall, Richard} discovered the advanced 
transform algorithm used by the GIMPS program. George Woltman implemented
Crandall's algorithm in machine language, thereby producing a prime-search 
program of unprecedented efficiency,
and that work led to the successful GIMPS project.

@@

% --------------------------------------------------------------------------
\section[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See also appendix \ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of this method within
    the tutorial ``World of Primes''.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

\vskip +25 pt
\paragraph*{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\section{Special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers \texorpdfstring{$f(n) = 2^{2^n} + 1$}{f(n) = 2\^{}2\^{}n + 1}]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g.\ \cite[p. 176]{pr:Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See chapter~\ref{primes:_Appendix_Sage-Samples} ``\nameref{primes:_Appendix_Sage-Samples}''
   for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See chapter~\ref{primes:_Appendix_Sage-Samples} ``\nameref{primes:_Appendix_Sage-Samples}''.).}.

@@

The above Catalan-Mersenne numbers are from $C_2$ a subset of the double Mersenne
primes\footnote{%
\url{http://en.wikipedia.org/wiki/Catalan%E2%80%93Mersenne_number}
          }.
A double Mersenne prime is a Mersenne prime of the form
$$ M_{M_p} = 2 ^ {2^p - 1} -1 $$
% $ M_{M_p} = 2 ^ {2^p - 1} -1 $ 
where p is a Mersenne prime exponent and $M_p$ ist a prime Mersenne number.

@@

$M_{M_p}$ is known to be prime for p = 2, 3, 5, 7, and there has the according values:
7, 127, 2.147.483.647, 170.141.183.460.469.231.731.687.303.715.884.105.727.\footnote{%
Sage\index{Sage} calculates this with:
\begin{Verbatim}
sage: for p in (2,3,5,7): Mp=(2^p)-1; MMp=(2^Mp)-1; print Mp, MMp;
....: 
3 7
7 127
31 2147483647
127 170141183460469231731687303715884105727
\end{Verbatim}
}

For p = 13, 17, 19, and 31, the corresponding double Mersenne numbers are not prime.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\section{Density and distribution of the primes}

@@

For $n > 5$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer prime numbers than even natural numbers but more prime numbers than
square numbers\footnote{%
Please refer to \hyperlink{ntePrimzahl}{the table \ref{s:ntePrimzahl}}
}.

>>>>

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Dec. 2012)

@@

% --------------------------------------------------------------------------
\section{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers (composite
numbers) can be formed by multiplying prime numbers.

\noindent The {\em natural} \index{Number!natural} numbers
$\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

\noindent Prime numbers are particularly important for three reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (e.g. elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorized in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1013 = 1 \cdot 1013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors not equal 1 are called 
\index{Number!composite} {\em composite} numbers. 
These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \url{http://primes.utm.edu/largest.html}
and at
     \url{http://primes.utm.edu/mersenne/index.html}.
}:

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2012)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

When the project started in 2002 there have been 17 possible candidates $< 78557$ (this is the reason for the project's name ``Seventeen or Bust''). It is sufficient to find one single counter-example, to exclude a candidate k, which means to find a single $n>=1$, where $N = k \cdot 2^{n}+1$ is prime. So it is only a byproduct of this task that this also generates new monster primes.
} 
\addtocounter{footnote}{1}
\footnotetext{%
Generalized Fermat number: $ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $
\index{Fermat!number!generalized}  } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. In 1903,
Frank Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{Factorization} (expression as a product of prime factors)\footnote{%
  Using CT1\index{CrypTool 1} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CT1 can factorize with the quadratic sieve (QS) on a single PC in a reasonable
  time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CT1. \\
  With CT2 you can use both the QS and the GNFS (general number field sieve),
  and CT2 can distribute the calculations on several computers.\\
  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{Factorization!factoring records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
Landon Curt Noll\index{Noll, Landon Curt} lists all known Mersenne primes
including its date of discovery and its value as number and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +10 pt
For the first 41            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 47th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely.\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }

@@

Table \ref{Primes_GIMPS-Primes_table-reference} contains these Mersenne record
primes.\footnote{%
An up-to-date version can be found in the internet at
     \url{http://www.mersenne.org/history.htm}.
}$^,$\footnote{%
Always, when a new record is published in the respective forums the same
and often ironic discussions start: Does this kind of research have a deeper
sense? Can this result be applied for anything useful?
The answer is, that we don't know it yet. In fundamental research one
cannot see at once whether and how it brings mankind forward.
%But this is the normal case with basic research, that one cannot see at once, 
}

@@

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of July 2012)}
\label{Primes_GIMPS-Primes_table-reference}
\end{center}
\end{table}

@@

Richard Crandall\index{Crandall, Richard} discovered the advanced 
transform algorithm used by the GIMPS program. George Woltman implemented
Crandall's algorithm in machine language, thereby producing a prime-search 
program of unprecedented efficiency,
and that work led to the successful GIMPS project.

@@

% --------------------------------------------------------------------------
\section[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See NT learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See also appendix \ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of these methods within
    the tutorial ``{\bf World of Primes}''.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

\vskip +25 pt
\paragraph*{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: It is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +30 pt
% caused a warning black rectangle:
%\section{Special number types and the search for a formula for primes}
\section{Special types of numbers and the search for a formula for primes}
\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers \texorpdfstring{$f(n) = 2^{2^n} + 1$}{f(n) = 2\^{}2\^{}n + 1}]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides
    $f(5)$.\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g.\ \cite[p. 176]{pr:Scheid1994})
    } 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function delivers prime numbers for all values from $n=0$ to $n=79$
   .\footnote{%
   See chapter~\ref{primes:_Appendix_Sage-Samples},
   ``\nameref{primes:_Appendix_Sage-Samples}'' for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See chapter~\ref{primes:_Appendix_Sage-Samples}, ``\nameref{primes:_Appendix_Sage-Samples}''.).}.

@@

The above Catalan-Mersenne numbers are from $C_2$ a subset of the double
Mersenne primes.\footnote{%
\url{http://en.wikipedia.org/wiki/Catalan%E2%80%93Mersenne_number}
          }
A double Mersenne prime is a Mersenne prime of the form
$$ M_{M_p} = 2 ^ {2^p - 1} -1 $$
% $ M_{M_p} = 2 ^ {2^p - 1} -1 $ 
where p is a Mersenne prime exponent and $M_p$ ist a prime Mersenne number.

@@

$M_{M_p}$ is known to be prime for p = 2, 3, 5, 7, and there has the according values:
7, 127, 2.147.483.647, 170.141.183.460.469.231.731.687.303.715.884.105.727.\footnote{%
Sage\index{Sage} calculates this with:
\begin{Verbatim}
sage: for p in (2,3,5,7): Mp=(2^p)-1; MMp=(2^Mp)-1; B=is_prime(MMp); print p, Mp, MMp, B;
....: 
2  3 7  True
3  7 127  True
5  31 2147483647  True
7  127 170141183460469231731687303715884105727  True
\end{Verbatim}
}

For p = 11, 13, 17, 19, and 31, the corresponding double Mersenne numbers are not prime.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{h_Primes_Distrib-of-Primes}{}
\section{Density and distribution of the primes}
\label{l_Primes_Distrib-of-Primes}

@@

For $n > 5$, $P(n)$  lies between  $2n$  and  $n^2$. This means that there are
fewer prime numbers than even natural numbers, but more prime numbers than
square numbers.\footnote{%
Please refer to \hyperlink{ntePrimzahl}{the table \ref{s:ntePrimzahl}}
}
